### 2024-03-26 docs/projects/geography/chicago-bikes/index.qmd
Chicago bikes - rerendered

## Data sources:
- [Cyclistic bike share - Kaggle](https://www.kaggle.com/datasets/evangower/cyclistic-bike-share)
- [Cyclistic bike share - 2023 update - Kaggle](https://www.kaggle.com/datasets/godofoutcasts/cyclistic-bike-share-2023)
- [Chicago neighborhoods geojson by @blackmad](https://github.com/blackmad/neighborhoods/blob/master/chicago.geojson)

@@

::: {.sourceClojure}
```clojure
(ns index
  (:require [tablecloth.api :as tc]
            [clojure.math :as math]
            [tech.v3.datatype.datetime :as datetime]
            [clojure.string :as str]
            [scicloj.noj.v1.vis.hanami :as hanami]
            [aerial.hanami.templates :as ht]
            [fastmath-clustering.core :as clustering]
            [geo
             [geohash :as geohash]
             [jts :as jts]
             [spatial :as spatial]
             [io :as geoio]
             [crs :as crs]]
            [scicloj.kindly.v4.kind :as kind]
            [scicloj.kindly.v4.api :as kindly])
  (:import (org.locationtech.jts.geom Geometry Point Polygon Coordinate)
           (org.locationtech.jts.geom.prep PreparedGeometry
                                           PreparedLineString
                                           PreparedPolygon
                                           PreparedGeometryFactory)))
```
:::

@@

For metric computations we need to convert them to a coordinate system
which is locally correct in terms of distances in a region around Chicago:
NAD83 / Illinois East
[EPSG:26971](https://epsg.io/26971-1742)

@@

::: {.sourceClojure}
```clojure
(defn lat-lng->local-coords [lat lng]
  (-> (geo.jts/coordinate -89.27 37.06)
      geo.jts/point
      wgs84->Chicago
      geo.jts/coord
      ((fn [^Coordinate coord]
         [(.getX coord)
          (.getY coord)]))))
```
:::

@@

::: {.sourceClojure}
```clojure
(delay
  (lat-lng->local-coords -89.27 37.06))
```
:::

::: {.printedClojure}
```clojure
[216692.34803057025 44059.50658557552]

@@

::: {.sourceClojure}
```clojure
(def coord-colnames
  [:start_lat :start_lng :end_lat :end_lng])
```
:::

::: {.sourceClojure}
```clojure
(def processed-trips
  (-> raw-trips
      (tc/add-column :hour (fn [trips]
                             (->> trips
                                  :started_at
                                  (datetime/long-temporal-field :hours))))
      (tc/select-rows (fn [row]
                        (->> coord-colnames
                             (map row)
                             (every? some?))))
      (tc/map-columns :start-local-coords [:start_lat :start_lng] lat-lng->local-coords)
      (tc/map-columns :end-local-coords [:end_lat :end_lng] lat-lng->local-coords)
      (tc/map-columns :start-local-x [:start-local-coords] first)
      (tc/map-columns :start-local-y [:start-local-coords] second)
      (tc/map-columns :end-local-x [:end-local-coords] first)
      (tc/map-columns :end-local-y [:end-local-coords] second)))
```
:::

@@

TODO

@@

```{=html}
<div><script>vegaEmbed(document.currentScript.parentElement, {"usermeta":{"embedOptions":{"renderer":"svg"}},"width":400,"background":"floralwhite","title":{"text":"Chicago bike trips"},"layer":[{"data":{"url":"notebooks\/data\/chicago.geojson","format":{"type":"topojson"}},"mark":{"type":"geoshape","filled":false,"clip":true,"opacity":0.3}},{"encoding":{"longitude":{"field":"start_lng","type":"quantitative"},"opacity":{"value":0.1},"latitude":{"field":"start_lat","type":"quantitative"},"longitude2":{"field":"end_lng","type":"quantitative"},"latitude2":{"field":"end_lat","type":"quantitative"}},"usermeta":{"embedOptions":{"renderer":"svg"}},"mark":"rule","width":400,"background":"floralwhite","projection":{"type":"mercator"},"height":300}],"height":300,"data":{"url":"index_files\/1.csv","format":{"type":"csv"}}});</script></div>
```

@@

::: {.sourceClojure}
```clojure
(def clustering
  (-> processed-trips
      (tc/select-columns coord-colnames)
      tc/rows
      (clustering/k-means 100)
      (dissoc :data)))
```
:::

>>>>

@@

::: {.sourceClojure}
```clojure
(ns index
  (:require [tablecloth.api :as tc]
            [clojure.math :as math]
            [tech.v3.datatype.datetime :as datetime]
            [tech.v3.datatype.functional :as fun]
            [clojure.string :as str]
            [scicloj.noj.v1.vis.hanami :as hanami]
            [aerial.hanami.templates :as ht]
            [fastmath-clustering.core :as clustering]
            [geo
             [geohash :as geohash]
             [jts :as jts]
             [spatial :as spatial]
             [io :as geoio]
             [crs :as crs]]
            [scicloj.kindly.v4.kind :as kind]
            [scicloj.kindly.v4.api :as kindly])
  (:import (org.locationtech.jts.geom Geometry Point Polygon Coordinate)
           (org.locationtech.jts.geom.prep PreparedGeometry
                                           PreparedLineString
                                           PreparedPolygon
                                           PreparedGeometryFactory)))
```
:::

In this notebook, we will explore the data of a bike sharing company in Chicago.

The data provides information of start & end location & time for each bike trip.
It has been used in academic papers and blog posts discussing various aspects of
people's movement patterns in their everyday lives.

Our goal here is to recognize and visualize a few movement patterns.

## Background

### Various openly shared datasets:
- [Divvy Trips](https://data.cityofchicago.org/Transportation/Divvy-Trips/fg6s-gzvg/about_data) - Chicago data portal
- [Chicago Divvy Bicycle Sharing Data](https://www.kaggle.com/datasets/yingwurenjian/chicago-divvy-bicycle-sharing-data) - Kaggle
- [Cyclistic bike share](https://www.kaggle.com/datasets/evangower/cyclistic-bike-share) - Kaggle
- [Cyclistic bike share - 2023 update](https://www.kaggle.com/datasets/godofoutcasts/cyclistic-bike-share-2023) - Kaggle

### Research papers and blog posts
- [Understanding Spatiotemporal Patterns of Biking Behavior by Analyzing Massive Bike Sharing Data in Chicago](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0137922) - Zhou 2015
- [Bicycle-Sharing System Analysis and Trip](https://www.cs.uic.edu/~jzhang2/files/2016_mdm_paper.pdf) - Zhang el al. 2016
- [Spatial Cluster-Based Model for Static Rebalancing Bike Sharing Problem](https://www.mdpi.com/2071-1050/11/11/3205) - Lahoorpoor et al. 2019
- [Examining spatiotemporal changing patterns of bike-sharing usage during COVID-19 pandemic](https://www.sciencedirect.com/science/article/pii/S0966692321000508) - Hu et al. 2021
- [Analysis: Chicago’s Divvy Bike Share](https://andrewluyt.github.io/divvy-bikeshare/analysis-report.html) - Luyt 2021

## Data sources for this notebook

1. [Chicago neighborhoods geojson](https://github.com/blackmad/neighborhoods/blob/master/chicago.geojson) - part of [a great collection](https://github.com/blackmad/neighborhoods/) by [@blackmad](https://github.com/blackmad).
We keep this file under "notebooks/data/". By default, [Clay](https://scicloj.github.io/clay) pushed data files under "notebooks" to the target render directory.
This will allow us to consume the file from the browser for data visualizations.

2. [Cyclistic bike share - 2023 update](https://www.kaggle.com/datasets/godofoutcasts/cyclistic-bike-share-2023) - shared on Kaggle by Jayprakash Kumar.
This dataset contains separate data files for different months. We will use only one of them here (Apr. 2023) and keep it under "data/".

@@

For metric computations (e.g., k-means clustering),
we need to convert them to a coordinate system
which is locally correct in terms of distances in a region around Chicago:
NAD83 / Illinois East
[EPSG:26971](https://epsg.io/26971)

@@

::: {.sourceClojure}
```clojure
(defn lng-lat->local-coords [lng lat]
  (-> (geo.jts/coordinate lng lat)
      geo.jts/point
      wgs84->Chicago
      geo.jts/coord
      ((fn [^Coordinate coord]
         [(.getX coord)
          (.getY coord)]))))
```
:::

@@

::: {.sourceClojure}
```clojure
(delay
  [(lng-lat->local-coords -89.27 37.06)
   (lng-lat->local-coords -87.02 42.5)])
```
:::

::: {.printedClojure}
```clojure
[[216692.34803057025 44059.50658557552]
 [407952.845339763 648476.2402163395]]

@@

::: {.sourceClojure}
```clojure
(def processed-trips
  (-> raw-trips
      ;; Compute starting hours:
      (tc/add-column :hour (fn [trips]
                             (->> trips
                                  :started_at
                                  (datetime/long-temporal-field :hours))))
      ;; Make sure the latitude and longitude are nonmissing:
      (tc/select-rows (fn [row]
                        (->> [:start_lat :start_lng :end_lat :end_lng]
                             (map row)
                             (every? some?))))
      ;; Add local Chicago coordinates:
      (tc/map-columns :start-local-coords [:start_lng :start_lat] lng-lat->local-coords)
      (tc/map-columns :end-local-coords [:end_lng :end_lat] lng-lat->local-coords)
      (tc/map-columns :start-local-x [:start-local-coords] first)
      (tc/map-columns :start-local-y [:start-local-coords] second)
      (tc/map-columns :end-local-x [:end-local-coords] first)
      (tc/map-columns :end-local-y [:end-local-coords] second)))
```
:::

@@

::: {.sourceClojure}
```clojure
(-> processed-trips
    (tc/random 1000 {:seed 1})
    (tc/add-column :local-L2 #(fun/sqrt
                               (fun/+ (fun/sq (fun/- (:start-local-x %)
                                                     (:end-local-x %)))
                                      (fun/sq (fun/- (:start-local-y %)
                                                     (:end-local-y %))))))
    (tc/add-column :global-L2 #(fun/sqrt
                                (fun/+ (fun/sq (fun/- (:start_lat %)
                                                      (:end_lat %)))
                                       (fun/sq (fun/- (:start_lng %)
                                                      (:end_lng %))))))
    (hanami/plot ht/point-chart
                 {:X :local-L2
                  :Y :global-L2}))
```
:::

```{=html}
<div><script>vegaEmbed(document.currentScript.parentElement, {"encoding":{"y":{"field":"global-L2","type":"quantitative"},"x":{"field":"local-L2","type":"quantitative"}},"usermeta":{"embedOptions":{"renderer":"svg"}},"mark":{"type":"circle","tooltip":true},"width":400,"background":"floralwhite","height":300,"data":{"url":"index_files\/0.csv","format":{"type":"csv"}}});</script></div>
```

The `local-L2` quantity, compted from the local Chicago coordinates,
is a decent approximation of actual distance.
We see that the `global-L2` quantity, computed accordingly
from latitude and longitude, is not directly related to
this quantity and might be misleading for metric tasks
such as clustering.

@@

```{=html}
<div><script>vegaEmbed(document.currentScript.parentElement, {"usermeta":{"embedOptions":{"renderer":"svg"}},"width":400,"background":"floralwhite","title":{"text":"Chicago bike trips"},"layer":[{"data":{"url":"notebooks\/data\/chicago.geojson","format":{"type":"topojson"}},"mark":{"type":"geoshape","filled":false,"clip":true,"opacity":0.3}},{"encoding":{"longitude":{"field":"start_lng","type":"quantitative"},"opacity":{"value":0.1},"latitude":{"field":"start_lat","type":"quantitative"},"longitude2":{"field":"end_lng","type":"quantitative"},"latitude2":{"field":"end_lat","type":"quantitative"}},"usermeta":{"embedOptions":{"renderer":"svg"}},"mark":"rule","width":400,"background":"floralwhite","projection":{"type":"mercator"},"height":300}],"height":300,"data":{"url":"index_files\/2.csv","format":{"type":"csv"}}});</script></div>
```

@@

::: {.sourceClojure}
```clojure
(def clustering
  (-> processed-trips
      (tc/select-columns [:start-local-x
                          :start-local-y
                          :end-local-x
                          :end-local-y])
      tc/rows
      (clustering/k-means 100)
      (dissoc :data)))
```
:::
